Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2013-11-19 paper/paper.tex
Rewrote the second paragraph, which is hopefully more focused now. Fixed some grammar issues and edited a paragraph in methods.

\newcommand{\fixme}[1]{\red{[#1]}}
\newcommand{\davidsays}[1]{{\color{red} [\green{David:} \emph{#1}]}}

@@

\author{Jeff B. Schulte}
\affiliation{Department of Physics, Oregon State University}
\author{Rene WHAT. Zeto}
\affiliation{Department of Physics, Oregon State University}
\author{David Roundy}
\affiliation{Department of Physics, Oregon State University}

@@

\section{Introduction}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  During this process a long FtsZ polymer chain develops on
the cell wall in the center region of the cell that dictates the plane
of division~\cite{adams2009bacterial,
  lutkenhaus2007assembly}. Previous experimental studies have shown
that the MinC protein, known to inhibit the FtZ
polymer\cite{shen2010examination}, exhibits a pole to pole oscillatory
behavoir in conjunction with the MinD and MinE proteins, while the
oscillating MinE tends to be more localized in the center of the
cell~\cite{hu1999topological, fu2001mine, shapiro2009and, yu1999ftsz,
  raskin1999rapid}. The MinC protein will then have a higher time
averaged concentration in the cell poles as opposed to the center
region of the cell, aiding in prohibiting the FtZ from developing in
the wrong region.

It is vital for the mechanism by which the cell accurately divides to
be robust and indeed it has been shown to
be\cite{touhami2006temperature}.  Studies have shown that the
oscillations can effectively find poles in abberent cell
shapes~\cite{corbin2002exploring} \cite{juarez2010changes}.  Varma et
al. have studied three pronged connected tube shapes, both
experimentally and in simulation\cite{varma2008min} and have shown
that even in these oscillations develop in regular patterns. However,
Mannik et al. have shown that there are limits to this
robustness. Forcing cells into flattened, irregular cell shapes
adversly effects the Min system's ability to maintain their regular
oscillatory behavior \fixme{this link doesn't work =
  cite{mannik2012robustness}} \cite{mannik2010bacteria}
\cite{mannik2009bacterial}.  We follow the experimental work of Mannik
et and use Huang's mean field differential equation reaction model to
explore the model's ability to find regular oscillatory behavoir in a
series of abberent shapes.  \davidsays{This paragraph could use some
  work, and should reference Mannik early, and emphasize his research.
  First paragraph sets the stage for why our work is important by
  introducing the MinD system.  Second paragraph should explain why we
  are studying the Min system in funny cell shapes.}

A significant amount of work has been done to develop protein reaction
and diffusion models that exhibit accurate macroscopic dynamics of the
MinD protein system. Early models involved free proteins that affect
eachothers' rates of diffusion and membrane attachement but do not
combine into compound states~\cite{meinhardt2001pattern}.  In 2003
Huang improved upon this work with a simple model based on MinD-MinE
combination, ATPase hydrolysis, and MinD membrane attechent that when
simulated exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is more
likely to attach to the membrane when MinD is already clustered there
(following observed non-linear attachement of minD on the cell
membrane), and is stationary once attached.  A number of studies have
used an approach similar in that they do not rely on the ability of
MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Variations of the Huang 2003 model that
stochasticly account for variations of molecular interaction
\cite{fange2006noise} and as well monte-carlo simulations that
implement stochastic version of Huang's mean filed reaction rates
confirm the major results obtained by Huang's model and more
successfully predict experimentally observed oscillations in round
cell phenotypes~\cite{drew2005polymerization, fange2006noise,
  huang2004min}. Biochemical models of broader scope have also been
used to study the MinD system and show consistent
results.\cite{arjunan2010new}.  However, in general the results of the
stochastic and monte-carlo simulations are similar to those given by
Huang's mean field results.

@@

\begin{multline}
  \frac{\partial \rho_{D:ADP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ADP}-k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  +\delta(d_w)k_{de}\sigma_{DE}\hspace{3.4cm}
\end{multline}
\begin{multline}
  \frac{\partial \rho_{D:ATP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ATP}+k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  -\delta(d_w)[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
\end{multline}
\begin{multline}
  \frac{\partial \rho_E}{\partial t} = \mathcal{D}_E\nabla^2\rho_E+\delta(d_w)k_{de}\sigma_{DE}
  -\delta(d_w)k_E \sigma_D \rho_E
\end{multline}
\begin{multline}
  \frac{\partial \sigma_d}{\partial t} = -k_E\rho_d\rho_E
  +[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
  \label{eq:d-on-wall}
\end{multline}
\begin{multline}
  \frac{\partial \sigma_{DE}}{\partial t} = -k_{de}\sigma_{DE}+k_E\sigma_D\rho_E\hspace{3cm}
  \label{eq:FifthPDE}
\end{multline}

Where $\rho$ is cytoplasmic protein density ($\mu m^{-3}$), $\sigma$
is membrane bound density ($\mu m^{-2}$), $\mathcal{D}_D$ and
$\mathcal{D}_{E}$ th diffusions constants for MinD and MinE,
respectively, $k_D^{\textrm{ADP $\rightarrow$ ATP}}$ the rate of
conversion from MinD:ADP to the MinD:ATP complex, $k_D$ the rate of
MinD:ATP attachement to the membrane when no protein is already
attached there, $k_{dD}$ the increase of this rate when MinD:ATP is
present on the membrane, $k_{de}$ the rate of hydrolisis of the
MinD:MinE:ATP complex, $k_E$ the rate of cytoplasmic MinE binding to
membrane bound MinD:ATP complex, and $d_w$ is the distance from the
point in space to the closest wall (which will always be
perpendicular to this distance).  $\delta(d_w)$ has units of
$\mu m^{-1}$ and will be zero everywhere except at the wall.
Equations \ref{eq:d-on-wall} and \ref{eq:FifthPDE} will only be
relevent at the membrane because the membrane bound density values
will have no meaning in the cytoplasm where there is no membrane.

@@

A 3d grid is constructed in cartesian coordinates with a grid spacing
of .05\micron. We define our cell shapes and solve the
reaction-diffusion equations numerically to observe the time evolution
of the MinD and MinE concentrations inside the cell.

We start our cells with the MinD and MinE concentrations that are
reported as wild type concentrations by Huang et al . We as well use the same reaction
diffusion constants and and reaction rates,

>>>>

\newcommand{\fixme}[1]{\red{[#1]}}
\newcommand{\davidsays}[1]{{\color{red} [\green{David:} \emph{#1}]}}
\newcommand{\renesays}[1]{{\color{red} [\green{Rene:} \emph{#1}]}}

@@

\author{Jeff B. Schulte}
\affiliation{Department of Physics, Oregon State University}
\author{Rene W. Zeto}
\affiliation{Department of Physics, Oregon State University}
\author{David Roundy}
\affiliation{Department of Physics, Oregon State University}

@@

\section{Introduction}
It is vital that during the process of bacterial cell division, a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  A long FtsZ polymer chain develops on the cell wall, in the
center region of the cell, the region which dictates the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer\cite{shen2010examination}, exhibits pole to
pole oscillatory behavior in conjunction with the MinD and MinE
proteins, while the oscillating MinE tends to be more localized in the
center of the cell~\cite{hu1999topological, fu2001mine,
  shapiro2009and, yu1999ftsz, raskin1999rapid}. The MinC protein will
then have a higher time averaged concentration in the cell poles, as
opposed to the center region of the cell, aiding in prohibiting the
FtZ from developing in the wrong region.

The cell division process must be able to prevent minicelling, even
under irregular conditions. Studies have shown that it is capable of
performing this task for fairly perturbed cell
shapes\cite{touhami2006temperature}. Mannik et al. have done some
early experimental work showing regular oscillations of cell division
systems in bacteria that were physically forced into unusual
shapes. Varma et al. studied one particular example of a perturbed
shape, a three-pronged tube, and found both experimental and
simulation results\cite{varma2008min} supporting the idea that certain
systems inside the cell begin to undergo oscillations inside the
cell. These oscillations seem to seek out the extreme poles in the
cell shape\cite{corbin2002exploring} \cite{juarez2010changes}, and
effectively prevent minicelling in most cases. However, Mannik et
al. have also shown that there are limitations to this
mechanism\cite{mannik2010bacteria} \cite{mannik2009bacterial}. One
instance of this behavior is observable when the cell has been forced
into a flattened, irregular shape. Oscillations in this sort of cell
shape are far from the regular sorts of oscillations seen in the
computational work done by Huang et al\cite{huang2003dynamic}. We seek
to apply a computational model to explore the more extreme or
irregular cell shapes, in order to determine the limits of the steady
cell oscillations seen by Mannik et al.

%is this necessary? was in old paragraph.
%\fixme{this link doesn't work =
%  cite{mannik2012robustness}} \cite{mannik2010bacteria}
%\cite{mannik2009bacterial}

A significant amount of work has been done to develop protein reaction
and diffusion models that exhibit accurate macroscopic dynamics of the
MinD protein system. Early models involved free proteins that affect
each others' rates of diffusion and membrane attachment, but do not
combine into compound states~\cite{meinhardt2001pattern}.  In 2003
Huang improved upon this work with a simple model based on MinD-MinE
combination, ATPase hydrolysis, and MinD membrane attachment that when
simulated, exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is more
likely to attach to the membrane when MinD is already clustered there
(following observed non-linear attachement of minD on the cell
membrane), and is stationary once attached.  A number of studies have
used an approach similar in that they do not rely on the ability of
MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Variations of the Huang 2003 model that
stochasticly account for variations of molecular interaction
\cite{fange2006noise} and monte-carlo simulations that implement this
stochastic version of Huang's mean field reaction rates confirm the
major results obtained by Huang's model, and more successfully predict
experimentally observed oscillations in round cell
phenotypes~\cite{drew2005polymerization, fange2006noise,
  huang2004min}. Biochemical models of broader scope have also been
used to study the MinD system and show consistent
results.\cite{arjunan2010new}.  However, in general, the results of the
stochastic and monte-carlo simulations are similar to those given by
Huang's mean field results.

@@

\begin{multline}
  \frac{\partial \rho_{D:ADP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ADP}-k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  +\delta(d_w)k_{de}\sigma_{DE},\hspace{3.4cm}
\end{multline}
\begin{multline}
  \frac{\partial \rho_{D:ATP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ATP}+k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  -\delta(d_w)[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP},
\end{multline}
\begin{multline}
  \frac{\partial \rho_E}{\partial t} = \mathcal{D}_E\nabla^2\rho_E+\delta(d_w)k_{de}\sigma_{DE}
  -\delta(d_w)k_E \sigma_D \rho_E,
\end{multline}
\begin{multline}
  \frac{\partial \sigma_d}{\partial t} = -k_E\rho_d\rho_E
  +[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP},
  \label{eq:d-on-wall}
\end{multline}
\begin{multline}
  \frac{\partial \sigma_{DE}}{\partial t} = -k_{de}\sigma_{DE}+k_E\sigma_D\rho_E\hspace{3cm}
  \label{eq:FifthPDE},
\end{multline}

where $\rho$ is cytoplasmic protein density ($\mu m^{-3}$), $\sigma$
is membrane bound density ($\mu m^{-2}$), $\mathcal{D}_D$ and
$\mathcal{D}_{E}$ th diffusions constants for MinD and MinE,
respectively, $k_D^{\textrm{ADP $\rightarrow$ ATP}}$ the rate of
conversion from MinD:ADP to the MinD:ATP complex, $k_D$ the rate of
MinD:ATP attachement to the membrane when no protein is already
attached there, $k_{dD}$ the increase of this rate when MinD:ATP is
present on the membrane, $k_{de}$ the rate of hydrolisis of the
MinD:MinE:ATP complex, $k_E$ the rate of cytoplasmic MinE binding to
membrane bound MinD:ATP complex, and $d_w$ is the distance from the
point in space to the closest wall (which will always be
perpendicular to this distance).  $\delta(d_w)$ has units of
$\mu m^{-1}$ and will be zero everywhere except at the wall.
Equations \ref{eq:d-on-wall} and \ref{eq:FifthPDE} will only be
relevent at the membrane because the membrane bound density values
will have no meaning in the cytoplasm where there is no membrane.

@@

We start our cells with the MinD and MinE concentrations that are
reported as wild type concentrations by Huang et al. These
concentration values are 1273 $\mu m^{-3}$ and 445 $\mu m^{-3}$,
respectively. A 3d grid is then constructed in cartesian coordinates,
with a grid spacing of .05\micron. We define our cell shapes and solve
the reaction-diffusion equations numerically to observe the time
evolution of the MinD and MinE concentrations inside the
cell. \renesays{This last sentence sweeps a lot under the rug. How
  much more detail should be given to the computational method?}
